### 2024-07-08 basic-statistics.qmd
20240708 - robust indices of central tendency and variability

```{r}
library("petersenlab")
library("pwr")
library("pwrss")
library("WebPower")
library("grid")
library("tidyverse")
```

@@

- mean
- median
- mode

The mean of $X$ (written as: $\bar{X}$) is calculated as:

@@

That is, to compute the mean, sum all of the values and divide by the number of values ($n$).

The median is determined as the value at the 50th percentile (i.e., the value that is higher than 50% of the values and is lower than the other 50% of values).

The mode is the most common/frequent value.

@@

- standard deviation
- variance
- range
- minimum and maximum
- interquartile range (IQR)

The (sample) variance of $X$ (written as: $s^2$) is calculated as:

@@

The (sample) standard deviation of $X$ (written as: $s$) is calculated as:

@@

The range is calculated of $X$ is calculated as:

@@

The interquartile range (IQR) is calculated as:

@@

@@

#mydata %>% 
#  summarise(across(
#      everything(),
#      .fns = list(
#        n = ~ length(na.omit(.)),
#        missingness = ~ mean(is.na(.)) * 100,
#        M = ~ mean(., na.rm = TRUE),
#        SD = ~ sd(., na.rm = TRUE),
#        min = ~ min(., na.rm = TRUE),
#        max = ~ max(., na.rm = TRUE),
#        range = ~ max(., na.rm = TRUE) - min(., na.rm = TRUE),
#        IQR = ~ IQR(., na.rm = TRUE),
#        median = ~ median(., na.rm = TRUE),
#        mode = ~ petersenlab::Mode(., multipleModes = "mean"),
#        skewness = ~ psych::skew(., na.rm = TRUE),
#        kurtosis = ~ psych::kurtosi(., na.rm = TRUE)),
#      .names = "{.col}.{.fn}")) %>%
#    pivot_longer(
#      cols = everything(),
#      names_to = c("variable","index"),
#      names_sep = "\\.") %>% 
#    pivot_wider(
#      names_from = index,
#      values_from = value)
```

>>>>

```{r}
library("petersenlab")
library("DescTools")
library("pwr")
library("pwrss")
library("WebPower")
library("grid")
library("tidyverse")
```

@@

- mean
- median
- Hodges-Lehmann statistic (aka pseudomedian)
- mode

The mean of $X$ (written as: $\bar{X}$) is calculated as in @eq-IQR:

@@

That is, to compute the mean, sum all of the values and divide by the number of values ($n$).
One issue with the mean is that it is sensitive to extreme (outlying) values.
For instance, the mean of the values of 0, 0, 10, 15, 20, 30, and 1000 is `{r} exampleValues_mean`.

```{r}
#| code-fold: true

exampleValues <- c(0, 0, 10, 15, 20, 30, 1000)
exampleValues_mean <- mean(exampleValues)
```

The median is determined as the value at the 50th percentile (i.e., the value that is higher than 50% of the values and is lower than the other 50% of values).
Compared to the mean, the median is less influenced by outliers.
The median of the values of 0, 0, 10, 15, 20, 30, and 1000 is `{r} exampleValues_median`.

```{r}
#| code-fold: true

exampleValues_median <- median(exampleValues)
```

The Hodges-Lehmann statistic (aka pseudomedian) is computed as the median of all pairwise means, and it is also robust to outliers.
The pseudomedian of the values of 0, 0, 10, 15, 20, 30, and 1000 is `{r} exampleValues_pseudomedian`.

```{r}
#| code-fold: true

exampleValues_pseudomedian <- DescTools::HodgesLehmann(exampleValues)
```

The mode is the most common/frequent value.
The mode of the values of 0, 0, 10, 15, 20, 30, and 1000 is `{r} exampleValues_mode`.

```{r}
#| code-fold: true

exampleValues_mode <- petersenlab::Mode(exampleValues)
```

@@

- standard deviation
- variance
- range
- minimum and maximum
- interquartile range (IQR)
- median absolute deviation

The (sample) variance of $X$ (written as: $s^2$) is calculated as in @eq-variance:

@@

where $X_i$ is each data point, $\bar{X}$ is the mean of $X$, and $n$ is the number of data points.

The (sample) standard deviation of $X$ (written as: $s$) is calculated as in @eq-sd:

@@

The range is calculated of $X$ is calculated as in @eq-range:

@@

The interquartile range (IQR) is calculated as in @eq-IQR:

@@

The median absolute deviation (MAD) is the median of all deviations from the median, and is calculated as in @eq-medianAbsoluteDeviation:

$$
\text{MAD} = \text{median}(|X_i - \tilde{X}|)
$$ {#eq-medianAbsoluteDeviation}

where $\tilde{X}$ is the median of `X`.
Compared to the standard deviation, the median absolute deviation is more robust to outliers.

@@

#mydata %>% 
#  summarise(across(
#      everything(),
#      .fns = list(
#        n = ~ length(na.omit(.)),
#        missingness = ~ mean(is.na(.)) * 100,
#        M = ~ mean(., na.rm = TRUE),
#        SD = ~ sd(., na.rm = TRUE),
#        min = ~ min(., na.rm = TRUE),
#        max = ~ max(., na.rm = TRUE),
#        range = ~ max(., na.rm = TRUE) - min(., na.rm = TRUE),
#        IQR = ~ IQR(., na.rm = TRUE),
#        mad = ~ mad(., na.rm = TRUE),
#        median = ~ median(., na.rm = TRUE),
#        pseudomedian = ~ DescTools::HodgesLehmann(., na.rm = TRUE),
#        mode = ~ petersenlab::Mode(., multipleModes = "mean"),
#        skewness = ~ psych::skew(., na.rm = TRUE),
#        kurtosis = ~ psych::kurtosi(., na.rm = TRUE)),
#      .names = "{.col}.{.fn}")) %>%
#    pivot_longer(
#      cols = everything(),
#      names_to = c("variable","index"),
#      names_sep = "\\.") %>% 
#    pivot_wider(
#      names_from = index,
#      values_from = value)
```
